Computational methods in landscape ecology

### 2024-06-03 Sections/spatial-patterns.qmd
adds first draft of the  Operations on spatial patterns section

@nowosadMotifOpensourceTool2021

>>>>

Various metrics are often used to describe spatial patterns in landscape ecology.
<!-- limits, correlations?? -->
At the same time, the spatial patterns could be analyzed using various operations, such as comparing, searching, or grouping [@turnerSpatialTemporalAnalysis1990;@wickhamMappingAnalyzingLandscape1994; @jasiewiczGeoPATToolboxPatternbased2015; @remmelDistributionsHyperLocalConfiguration2020b;@nowosadMotifOpensourceTool2021].
The above operations are based on two main ideas: the spatial signatures and the dissimilarity measures.
A spatial signature is a multi-numerical representation of a landscape pattern for a given area.
It may be created based on a set of (preferably uncorrelated) landscape metrics, or be derived from the data itself, e.g., by using the co-occurrence matrix [@nowosadPatternbasedIdentificationMapping2021].
The dissimilarity measure is a function that quantifies the difference between two spatial signatures, with low values indicating similar patterns.

The most basic operation, the comparison of spatial patterns, could be performed for just a pair of areas or the same area in two different moments.
It can be also extended in a straightforward way by dividing the landscape into multiple smaller areas and comparing them for two moments in time.
The comparison of spatial patterns is often used to analyze the landscape dynamics, e.g., to detect the changes in the landscape structure over time [@duncanMonitoringSpatialPatterns2023].
The pattern-based search is a one to many operation, in which spatial signature of a query area is compared to the spatial signatures of all other areas.
This allows us to find areas with similar patterns to the query area, which could be used to identify areas with similar environmental conditions [@jasiewiczLandscapeSimilarityRetrieval2014].
Finally, the grouping of spatial patterns is a many to many operation, in which spatial signatures are calculated for all areas and then these areas are clustered based on their similarity [@nowosadPatternbasedIdentificationMapping2021].

<!-- motif -->
<!-- spquery -->
